Evaluation of Design Trade-offs for Adders\\ in Approximate Datapath

### 2014-11-19 WAP2015/wap2015.tex
add relevant work

\section{Introduction}
Circuit performance has been increased tremendously over the past decades. However, in recent years people have observed an end of performance improvement [xx] due to the scaling of CMOS technology to the nanometer regime. In order to boost circuit performance, two standard design techniques are commonly adopted: one is to heavily pipeline the datapath, the other is to reduce the design precision. For the first technique, pipelining can be used to increase operating frequency, whereas the overall computational latency will not be affected. Therefore the application of this technique is limited in many embedded applications, which are often designed with strict latency specifications. The second technique is common especially with FPGA technology, which embodies the flexibility to implement customized variable representation and is able to optimize the operand word-length across the entire datapath to meet a given performance and area constraint. However, the precision loss will inevitably introduce quantization errors into the design.

@@

In this paper, we provide detailed evaluations of design trade-offs for adders, which are key arithmetic primitives for a broad range of applications. In order to provide the optimum design choice for adders, we examine several design constrains and options, such as accuracy, operating frequency, silicon area, different types of adder structure, as well as different forms of computer arithmetic for adder implementation. In comparison to the conventional design scenario that timing closure is achieved by truncating the word-length of operands, we take a radical step forward by adopting an alternative design scenario, which is named ``overclockign scenario''. In this scenario, circuits are operated far beyond the conservation region with full precision, while timing violation can be tolerated. Previous work has shown that for many applications, it is preferable to follow the overclocking scenario because the worst case is triggered by certain input patterns and hence it happens rarely [xxx]. In addition to the adder structures that are created based on standard binary arithmetic, such as ripple carry adder and carry select adder, we also evaluate adder with ``online arithmetic'' [xxx], which employs different style of data representation to eliminate carry propagation in conventional adders. We demonstrate experimentally on FPGAs that for limited area budgets, applying the overclocking approach to ripple carry adders can be more beneficial than using fast adders to achieve similar latency. With more relaxed area constraints, the online adder outperforms in terms of accuracy or performance. Potentially our study will provide circuit designers with guidance and extra options when balancing between various design trade-offs.

In summary, the main contributions of this paper are:
\begin{enumerate}
  \item Detailed evaluation of three adder structures: ripple carry adder, carry select adder and online adder; 
  \item Evaluation of design scenarios by demonstrating the optimum design choice under the consideration of accuracy-performance-area trade-offs.
\end{enumerate}

@@

\section{Background}\label{sec:Background}
\subsection{Approximate Datapath Design}
Typically this can be achieved by either reducing the supply voltage [xxx] or increasing the operating frequency beyond the rated value [xxx].

@@

As an alternative form of computer arithmetic, online arithmetic has been used in numerous applications such as signal processing and control algorithms~\cite{Online_FPGADSP,Online_Control}. Online arithmetic was originally designed for digit-serial operation, as illustrated in Fig~\ref{Fig:OnlineDataFlow}. It can be seen that in order to generate the first output digit, $\delta$ digits of inputs are required, where $\delta$ is called the ``online delay''. $\delta$ is normally a small constant, which is independent of the precision. For ease of discussion, for the rest of this paper, the input data is assumed to be fixed point numbers in the range $(-1,1)$. Based on this premise, the online representation of $N$-digit operands and result at iteration $j$ are given by (\ref{Eq:Online_Operands}), where $j\in[-\delta,N-1]$ and $r$ denotes the radix \cite{Ercegovac_Book}.
%
\begin{figure}[htbp]
  \centering
  %\vspace{-2.5ex}
  \includegraphics[width=.42\textwidth]{./figures/OnlineArithmetic_DataFlow.eps}
  %\vspace{-1ex}
  \caption{Dataflow in digit-serial online arithmetic, in which both inputs and outputs are processed from the MSD to the LSD. $\delta$ denotes the online delay.}
  \vspace{-2ex}
  \label{Fig:OnlineDataFlow}
\end{figure}
%
\begin{eqnarray}\label{Eq:Online_Operands}
\footnotesize
  X_{[j]}=\sum_{i=1}^{j+\delta}x_ir^{-i},~Y_{[j]}=\sum_{i=1}^{j+\delta}y_ir^{-i},~Z_{[j]}=\sum_{i=1}^{j}z_ir^{-i}
\normalsize
\end{eqnarray}

@@

For an $n$-bit RCA, it is composed of $n$ serial-connected full adders (FA). Typically the maximum frequency of RCA is determined by the longest carry propagation. Under the assumption that the carry propagation delay of each FA is a constant value $\mu$, the critical path of the RCA is: $\mu_{RCA}=n\mu$. For the sampling period $T_S$, it follows that if $T_S\geqslant\mu_{RCA}$, correct result will always be sampled; otherwise if $T_S<\mu_{RCA}$, intermediate be sampled and potentially generating errors. In the previous work, the mean value of overclocking error is modeled as given in (\ref{Eq:MeanError_RCA}), where coefficient $b$ is given in (\ref{Eq:b}) and it determines the maximum length of error-free carry propagation in bits.
%
\begin{eqnarray}\label{Eq:MeanError_RCA}
    E_O=\left\{
        \begin{matrix}
            2^{-b}-2^{-n-1}, & \textrm{if $b\leq n$}\\
            0, & \textrm{otherwise}
        \end{matrix}
        \right.
\end{eqnarray}
%
\begin{eqnarray}\label{Eq:b}
    b:=\left\lceil \frac{T_S}{\mu} \right\rceil=\left\lceil \frac{1}{\mu\cdot f_S}\right\rceil
\end{eqnarray}

On the other hand, a specific timing target can be met by truncating the word-length of RCA while timing violation is not permitted. This will result in truncation error for most data.  The mean value of truncation error is also modeled as given in~(\ref{Eq:MeanTruncError_RCA}) [xxx], where parameters $k$ and $n$ denote the word-length of RCA before and after truncation, respectively. 
%
\begin{eqnarray}\label{Eq:MeanTruncError_RCA}
    E_O=\left\{
        \begin{matrix}
            2^{-n}-2^{-k}, & \textrm{if $n<k$}\\
            0, & \textrm{otherwise}
        \end{matrix}
        \right.
\end{eqnarray}

@@

%In this paper, we take CSA with different stages into comparison when considering a variety of accuracy, performance and area trade-offs. The results are shown in Section.xxx.

\subsection{Online Adder}\label{subsec:online_adder}
%\subsubsection{Structure}
In addition to standard binary arithmetic, alternative forms of computer arithmetic was studied to boost performance. For example, adder with online arithmetic is also designed for low latency at the cost of more silicon area. The structure of a digit-parallel online adder (OA) where all signals represented with SD numbers of digit set $\{-1,0,1\}$ is shown in Fig.~\ref{Fig:Radix2SD_adder}. The module ``3:2'' denotes a 3:2 compressor, which takes three inputs and generates two outputs, and is logically equivalent to a full adder (FA). A major advantage of the redundant number system over the standard ripple-carry based arithmetic is that the propagation of carry is eliminated, resulting in a precision-independent computation time for addition. As labelled in Fig.~\ref{Fig:Radix2SD_adder}, ideally the computation delay of this adder is only two FA delays for any operand word-length, at the expenses of one extra FA for each digit of operands. 
%
\begin{figure}[tbp]
  \centering
  \includegraphics[width=.45\textwidth]{./figures/SDAdder.eps}
  \caption{An $N$-digit binary digit-parallel online adder. Both inputs and outputs are represented using SD representation. ``3:2'' denotes a 3:2 compressor.}
  \label{Fig:Radix2SD_adder}
\end{figure}

@@

However if the area budget is released to 55 LUTs, OA can be implemented with original word-length. Additionally CSA can be implemented with larger precision, but still not the full precision. Therefore as shown in Fig.~\ref{subfig:error_area_lut55}, OA serves as the optimum design choice with respect to a wide range of frequencies. For higher frequencies, the MRE of OA increases rapidly, and CSA with 4 stages outperforms for higher frequencies. Similar to Fig.~\ref{subfig:error_area_lut35}, the overclocked RCA is the optimum design choice for even higher frequencies, because in this case the CSA can only be implemented with small precisions, and the multiplexer delay limits the benefits of parallelism.

The results shown in Fig.~\ref{Fig:Error_Area} can be generalized with a variety of area constraints. In this case, the optimum design methods with respect to different operating frequencies and area consumptions are illustrated in Fig.~\ref{Fig:adder_3d_FreqArea}. 
%
\begin{figure*}[tbp]
  \centering
  \includegraphics[width=.8\textwidth]{./figures/exp/3d_FreqArea.eps}
  \caption{The maximum word-length of different adder structures.}
  \label{Fig:adder_3d_FreqArea}
\end{figure*}

\subsection{Optimum Adder Design Metric with Given Accuracy and Area Requirements}
%
\begin{figure*}[tbp]
  \centering
  \includegraphics[width=.8\textwidth]{./figures/exp/3d_ErrorArea.eps}
  \caption{The maximum word-length of different adder structures.}
  \label{Fig:adder_3d_ErrorArea}
\end{figure*}

>>>>

\section{Introduction}
Circuit performance has been increased tremendously over the past decades. However, in recent years people have observed an end of performance improvement~\cite{DarkSilicon2011} due to the scaling of CMOS technology to the nanometer regime. In order to boost circuit performance, two standard design techniques are commonly adopted: one is to heavily pipeline the datapath, the other is to reduce the design precision. For the first technique, pipelining can be used to increase operating frequency, whereas the overall computational latency will not be affected. Therefore the application of this technique is limited in many embedded applications, which are often designed with strict latency specifications. The second technique is common especially with FPGA technology, which embodies the flexibility to implement customized variable representation and is able to optimize the operand word-length across the entire datapath to meet a given performance and area constraint. However, the precision loss will inevitably introduce quantization errors into the design.

@@

In this paper, we provide detailed evaluations of design trade-offs for adders, which are key arithmetic primitives for a broad range of applications. In order to provide the optimum design choice for adders, we examine several design constrains and options, such as accuracy, operating frequency, silicon area, different types of adder structure, as well as different forms of computer arithmetic for adder implementation. In comparison to the conventional design scenario that timing closure is achieved by truncating the word-length of operands, we take a radical step forward by adopting an alternative design scenario, which is named ``overclockign scenario''. In this scenario, circuits are operated far beyond the conservation region with full precision, while timing violation can be tolerated. Previous work has shown that for many applications, it is preferable to follow the overclocking scenario because the worst case is triggered by certain input patterns and hence it happens rarely~\cite{SKfccm13}. In addition to the adder structures that are created based on standard binary arithmetic, such as ripple carry adder and carry select adder, we also evaluate adder with ``online arithmetic''~\cite{Ercegovac_OnlineOverview}, which employs different style of data representation to eliminate carry propagation in conventional adders. We demonstrate experimentally on FPGAs that for limited area budgets, applying the overclocking approach to ripple carry adders can be more beneficial than using fast adders to achieve similar latency. With more relaxed area constraints, the online adder outperforms in terms of accuracy or performance. Potentially our study will provide circuit designers with guidance and extra options when balancing between various design trade-offs.

In summary, the main contributions of this paper are:
\begin{enumerate}
  \item Detailed evaluation of three adder structures: ripple carry adder, carry select adder and online adder; 
  \item Evaluation of design scenarios and identification of the optimum design choice under the consideration of accuracy-performance-area trade-offs.
\end{enumerate}

@@

\section{Background}\label{sec:Background}
\subsection{Approximate Datapath Design}
The International Technology Roadmap for Semiconductors (ITRS) has identified that additional benefits of manufacturing, test, power and timing can be achieved by releasing the requirement of absolute correctness~\cite{semiconductor2007international}. A series of studies in this area called ``Better Than Worst-Case Design'' focused on creating circuits by operating beyond the safety margins. Typically this can be achieved by either reducing the supply voltage~\cite{NonUniformScaling} or increasing the operating frequency beyond the rated value~\cite{SKfccm13,JoshFCCM2012}. The Razor project~\cite{Razor2004} served as one of the exemplary designs, in which both supply voltage and clock frequency were scaled over the most conservative value, while the error rate was monitored by using a self-checking circuit. It was shown that over $30\%$ power saving can be obtained at the cost of around $1\%$ error rates.

Another stream of research was focus on proposing either approximate circuit structures or design tools by trading accuracy to performance and/or energy benefits. Work in this area was inspired by the fact that many applications, e.g. image processing and machine learning, were naturally error resilient. For example, code analysis tools~\cite{Truffle2012Uwash,EnerJ2011Uwash} were developed to divide the program into precise parts and approximate parts, which can then be mapped to hardware components with different speed-grades and supply voltages. As for the approximate hardware, numerous forms of architectures were designed for basic arithmetic operators such as adders [xxx] and multipliers [xxx]. For example, a under-designed multiplier unit was proposed, of which the worst case was replaced by a normal case based on straightforward Karnaugh-Map analysis~\cite{Undersigned2x2multiplier}. However, many of these techniques are designed at gate-level or transistor-level, and they cannot be directly applied onto existing hardware platforms such as FPGAs.

In this paper, we overcome the previous limitations by comparing three adder structures: ripple carry adder and carry select adder, which are widely employed in mainstream hardware platforms, together with online adder which can be also found when designing fast multipliers and embedded processors~\cite{RedundantMult_1987}. It is argued that our approach can be efficiently applied to existing designs, and we back up this hypothesis based on FPGA experiments.

@@

As an alternative form of computer arithmetic, online arithmetic has been used in numerous applications such as signal processing and control algorithms~\cite{Online_FPGADSP,Online_Control}. Online arithmetic was originally designed for digit-serial operation, as illustrated in Fig~\ref{Fig:OnlineDataFlow}. It can be seen that in order to generate the first output digit, $\delta$ digits of inputs are required, where $\delta$ is called the ``online delay''. $\delta$ is normally a small constant, which is independent of the precision. For ease of discussion, for the rest of this paper, the input data is assumed to be fixed point numbers in the range $(-1,1)$. Based on this premise, the online representation of $N$-digit operands and result at iteration $j$ are given by (\ref{Eq:Online_Operands}), where $j\in[-\delta,N-1]$ and $r$ denotes the radix \cite{Ercegovac_Book}.
%
\begin{figure}[tbp]
  \centering
  %\vspace{-2.5ex}
  \includegraphics[width=.42\textwidth]{./figures/OnlineArithmetic_DataFlow.eps}
  %\vspace{-1ex}
  \caption{Dataflow in digit-serial online arithmetic, in which both inputs and outputs are processed from the MSD to the LSD. $\delta$ denotes the online delay.}
  \vspace{-2ex}
  \label{Fig:OnlineDataFlow}
\end{figure}
%
\begin{eqnarray}\label{Eq:Online_Operands}
\footnotesize
  X_{[j]}=\sum_{i=1}^{j+\delta}x_ir^{-i},~Y_{[j]}=\sum_{i=1}^{j+\delta}y_ir^{-i},~Z_{[j]}=\sum_{i=1}^{j}z_ir^{-i}
\normalsize
\end{eqnarray}

@@

For an $n$-bit RCA, it is composed of $n$ serial-connected full adders (FA). Typically the maximum frequency of RCA is determined by the longest carry propagation. Under the assumption that the carry propagation delay of each FA is a constant value $\mu$, the critical path of the RCA is: $\mu_{RCA}=n\mu$. For the sampling period $T_S$, it follows that if $T_S\geqslant\mu_{RCA}$, correct result will always be sampled; otherwise if $T_S<\mu_{RCA}$, intermediate be sampled and potentially generating errors. In the previous work~\cite{SKfccm13}, the mean value of overclocking error is modeled as given in (\ref{Eq:MeanError_RCA}), where coefficient $b$ is given in (\ref{Eq:b}) and it determines the maximum length of error-free carry propagation in bits.
%
\begin{eqnarray}\label{Eq:MeanError_RCA}
    E_O=\left\{
        \begin{matrix}
            2^{-b}-2^{-n-1}, & \textrm{if $b\leq n$}\\
            0, & \textrm{otherwise}
        \end{matrix}
        \right.
\end{eqnarray}
%
\begin{eqnarray}\label{Eq:b}
    b:=\left\lceil \frac{T_S}{\mu} \right\rceil=\left\lceil \frac{1}{\mu\cdot f_S}\right\rceil
\end{eqnarray}

On the other hand, a specific timing target can be met by truncating the word-length of RCA while timing violation is not permitted. This will result in truncation error for most data.  The mean value of truncation error is also modeled as given in~(\ref{Eq:MeanTruncError_RCA})~\cite{SKfccm13}, where parameters $k$ and $n$ denote the word-length of RCA before and after truncation, respectively. 
%
\begin{eqnarray}\label{Eq:MeanTruncError_RCA}
    E_O=\left\{
        \begin{matrix}
            2^{-n}-2^{-k}, & \textrm{if $n<k$}\\
            0, & \textrm{otherwise}
        \end{matrix}
        \right.
\end{eqnarray}

@@

\subsection{Online Adder}\label{subsec:online_adder}
In addition to standard binary arithmetic, alternative forms of computer arithmetic was studied to boost performance. For example, adder with online arithmetic is also designed for low latency at the cost of more silicon area. The structure of a digit-parallel online adder (OA) where all signals represented with SD numbers of digit set $\{-1,0,1\}$ is shown in Fig.~\ref{Fig:Radix2SD_adder}. The module ``3:2'' denotes a 3:2 compressor, which takes three inputs and generates two outputs, and is logically equivalent to a full adder (FA). A major advantage of the redundant number system over the standard ripple-carry based arithmetic is that the propagation of carry is eliminated, resulting in a precision-independent computation time for addition. As labelled in Fig.~\ref{Fig:Radix2SD_adder}, ideally the computation delay of this adder is only two FA delays for any operand word-length, at the expenses of one extra FA for each digit of operands. 
%
\begin{figure}[tbp]
  \centering
  \includegraphics[width=.45\textwidth]{./figures/SDAdder.eps}
  \caption{An $N$-digit binary digit-parallel online adder. Both inputs and outputs are represented using SD representation. ``3:2'' denotes a 3:2 compressor.}
  \label{Fig:Radix2SD_adder}
\end{figure}

@@

However if the area budget is released to 55 LUTs, OA can be implemented with original word-length. Additionally CSA can be implemented with larger precision, but still not the full precision. Therefore as shown in Fig.~\ref{subfig:error_area_lut55}, OA serves as the optimum design choice with respect to a wide range of frequencies. For higher frequencies, the MRE of OA increases rapidly, and CSA with 4 stages outperforms for higher frequencies. Similar to Fig.~\ref{subfig:error_area_lut35}, the overclocked RCA is the optimum design choice for even higher frequencies, because in this case the CSA can only be implemented with small precisions, and the multiplexer delay limits the benefits of parallelism.
%
\begin{figure*}[tbp]
  \centering
  \includegraphics[width=.8\textwidth]{./figures/exp/3d_FreqArea.eps}
  \caption{The maximum word-length of different adder structures.}
  \label{Fig:adder_3d_FreqArea}
\end{figure*}

The results shown in Fig.~\ref{Fig:Error_Area} can be further extended by testing with a variety of area constraints. In this case, we can map the optimum design scenarios into the design space of different operating frequencies and area consumptions, as shown in Fig.~\ref{Fig:adder_3d_FreqArea}. It can be seen that in general the graph is divided into several regions. Firstly, if the frequency requirement is moderate whilst the area budget is large enough to implement an OA in full precision, it will be the optimum design. This is because long carry chains are eliminated from OA such that it can maintain full precision for a wide range of frequency values. Also this is in accordance with the results in Fig.~\ref{subfig:error_area_lut55}. Secondly, CSA is a better design choice for high frequency requirement and large area budget, because it is originally designed with high speed operation. Besides, each stage of CSA is implemented in the FPGA with fast carry logic [xxx], hence it runs even faster than OA. Thirdly, for a tighter area budget only part of OA and CSA can be implemented, whereas RCA still keeps original precision. In this situation, area becomes the limitation factor. Truncation errors occur for both OA and CSA and they are greater than the overclocking error of RCA. Therefore, the overclocked RCA is the optimum design option across almost the entire frequency domain. Furthermore, the precision of RCA is also limited for very stringent area constraints. This causes truncation errors for every design option. Nevertheless, RCA with either overclocking or truncation scenario equally serves to be optimal, because it maintains more precision than OA and CSA for a specific area budget.

\subsection{Optimum Adder Design Metric with Given Accuracy and Area Requirements}
%
\begin{figure*}[tbp]
  \centering
  \includegraphics[width=.8\textwidth]{./figures/exp/3d_ErrorArea.eps}
  \caption{.The original word-length for all adders is 32-bit.}
  \label{Fig:adder_3d_ErrorArea}
\end{figure*}
